Formalizing Stack Safety as a Security Property

### 2023-02-03 submission5/core.tex
ref to cref

In this section we introduce our security property definitions by means
of small code examples, using a simplified set of security-relevent operations for
calls, returns, and private allocations.
Figure \ref{fig:main} gives C code and corresponding compiled 32-bit RISC-V code
for a function {\tt main}, which
takes an argument {\tt secret} and initializes a local variable {\tt sensitive} to contain
potentially sensitive data.
Then {\tt main} calls another function {\tt f},
and afterward performs a test on {\tt sensitive} to decide whether
to output {\tt secret}.  Since {\tt sensitive} is initialized to 0,
the test should fail, and {\tt main} should instead output the return value of {\tt f}.
Output is performed by writing to the special global {\tt out},
and we assume that such writes are the only observable events in the system.

@@

We now consider how {\tt f} might misbehave and violate desirable
stack safety properties associated with {\tt main}. To put the violations in a
security framework, suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.
%
Leakage is most obviously viewed as a violation of {\tt main}'s {\it confidentiality}.
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (\(\wbcf\).)
%
In Figure \ref{subfig:WBCF2}, a different attack violates \(\wbcf\), this time
by returning to the correct program counter but with the wrong stack pointer.%
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which simplifies some later explanations.}

@@

The security semantics for this program is based
on the security-relevant events noted in the right columns of Figures~\ref{fig:main}
and~\ref{fig:f}, namely execution of instructions that allocate or deallocate space,
make a call, or make a return.

Our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
Figure \ref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The formal details of the security semantics are described in
\cref{sec:machine}, and the context evolution rules are formalized in \cref{fig:basicops}.)
Execution begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

@@

We now show how this security semantics can be used to define notions of confidentiality,
integrity, and correct control flow in such a way that many classes of
bad behavior, including the attacks in Figure~\ref{fig:f}, are
detected as security violations.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential. In Figure \ref{subfig:WBCF}, the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}.
In Figure \ref{subfig:WBCF2}, the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}
\end{center}

@@

Like \(\wbcf\), stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory in the function's
frame. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, any \(\sealed\) elements
are identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to \(\wbcf\), we define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43. As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

Consider the state \(\mach\) after step 5, whose context is \((V_2,\sigma)\), with the attacker from
Figure \ref{subfig:direct}. We take a variant state over the set of elements that are
\(\sealed\) in \(V_2\) (see Figure \ref{fig:variant2}.)
            [TODO: fix figure to avoid varying the entire stack.]
If we take a trace of execution from each state until it returns,
the traces may differ, in this case outputting 5 (the original value of {\tt secret}) and
4 (its value in the variant) respectively. This is a violation of
{\it internal confidentiality} (formalized in \cref{tab:props}, line 3a).

@@

But, in Figure \ref{subfig:indirect}, we also saw an attacker that exfiltrated the secret
by reading it and then returning it, in a context where the caller would output the returned
value. Figure \ref{fig:variant3} shows the behavior of the same variants under this attacker,
but in this case, there is no output during the call. Instead the value of {\tt secret} is
extracted and placed in {\tt a0}, the return value register. We wish to identify this as
a confidentiality violation, again by considering variants of the \(\sealed\)
elements in \(V_2\), but capturing the required property is subtle.

@@

\cref{sec:example} describes informally how the security context evolves as the system performs
security-relevant operations. Formally, we combine each machine state with a context
to create a {\it combined state} \(s = (\mach,\context)\) and lift the transition
to \(\stepstounder{}\) on combined states. 
At each step, the context updates based on the function
\(Op : \MACHS \rightarrow \CONTEXTS \rightarrow \psi \rightarrow \CONTEXTS\).
(Note that \(Op\) takes as its first argument the state {\it before} the step.)
Since a single step might correspond to multiple operations, we apply
\(Op\) as many times as needed, using \(\mathit{foldl}\).

@@

The mean-time-to-failure (MTTF) and average number of tests for various bugs can be found in
Table~\ref{tab:bug-table}, along with the average number of tests
it took to find the failure. Experiments were run in a desktop
machine equipped with i7-4790K CPU @ 4.0GHz with 32GB RAM.

>>>>

In this section we introduce our security property definitions by means
of small code examples, using a simplified set of security-relevent operations for
calls, returns, and private allocations.
\Cref{fig:main} gives C code and possible corresponding compiled 32-bit RISC-V code
for a function {\tt main}, which
takes an argument {\tt secret} and initializes a local variable {\tt sensitive} to contain
potentially sensitive data.
Then {\tt main} calls another function {\tt f},
and afterward performs a test on {\tt sensitive} to decide whether
to output {\tt secret}.  Since {\tt sensitive} is initialized to 0,
the test should fail, and {\tt main} should instead output the return value of {\tt f}.
Output is performed by writing to the special global {\tt out},
and we assume that such writes are the only observable events in the system.

@@

We now consider how {\tt f} might misbehave and violate desirable
stack safety properties associated with {\tt main}. To put the violations in a
security framework, suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in \cref{fig:f}.
%
Leakage is most obviously viewed as a violation of {\tt main}'s {\it confidentiality}.
In \cref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in \cref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(\cref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In \cref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (\(\wbcf\).)
%
In \cref{subfig:WBCF2}, a different attack violates \(\wbcf\), this time
by returning to the correct program counter but with the wrong stack pointer.%
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which keeps the step numbering uniform in~\cref{fig:exec1}.}

@@

The security semantics for this program is based
on the security-relevant events noted in the right columns of \cref{fig:main,fig:f},
namely execution of instructions that allocate or deallocate space,
make a call, or make a return.

Our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
\Cref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The formal details of the security semantics are described in
\cref{sec:machine}, and the context evolution rules are formalized in \cref{fig:basicops}.)
Execution begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

@@

We now show how this security semantics can be used to define notions of confidentiality,
integrity, and correct control flow in such a way that many classes of
bad behavior, including the attacks in \cref{fig:f}, are
detected as security violations.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential. In \cref{subfig:WBCF}, the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}.
In \cref{subfig:WBCF2}, the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}
\end{center}

@@

Like \(\wbcf\), stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory in the function's
frame. In \cref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, any \(\sealed\) elements
are identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in \cref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). \Cref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to \(\wbcf\), we define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43. As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

Consider the state \(\mach\) after step 5, whose context is \((V_2,\sigma)\), with the attacker from
\cref{subfig:direct}. We take a variant state over the set of elements that are
\(\sealed\) in \(V_2\) (see \cref{fig:variant2}.)
            [TODO: fix figure to avoid varying the entire stack.]
If we take a trace of execution from each state until it returns,
the traces may differ, in this case outputting 5 (the original value of {\tt secret}) and
4 (its value in the variant) respectively. This is a violation of
{\it internal confidentiality} (formalized in \cref{tab:props}, line 3a).

@@

But, in \cref{subfig:indirect}, we also saw an attacker that exfiltrated the secret
by reading it and then returning it, in a context where the caller would output the returned
value. \Cref{fig:variant3} shows the behavior of the same variants under this attacker,
but in this case, there is no output during the call. Instead the value of {\tt secret} is
extracted and placed in {\tt a0}, the return value register. We wish to identify this as
a confidentiality violation, again by considering variants of the \(\sealed\)
elements in \(V_2\), but capturing the required property is subtle.

@@

\Cref{sec:example} describes informally how the security context evolves as the system performs
security-relevant operations. Formally, we combine each machine state with a context
to create a {\it combined state} \(s = (\mach,\context)\) and lift the transition
to \(\stepstounder{}\) on combined states. 
At each step, the context updates based on the function
\(Op : \MACHS \rightarrow \CONTEXTS \rightarrow \psi \rightarrow \CONTEXTS\).
(Note that \(Op\) takes as its first argument the state {\it before} the step.)
Since a single step might correspond to multiple operations, we apply
\(Op\) as many times as needed, using \(\mathit{foldl}\).

@@

The mean-time-to-failure (MTTF) and average number of tests for various bugs can be found in
\cref{tab:bug-table}, along with the average number of tests
it took to find the failure. Experiments were run in a desktop
machine equipped with i7-4790K CPU @ 4.0GHz with 32GB RAM.
